---
title: "Data Cleaning Document"
output: pdf_document
author: "Archi Patel,Amanda Contreras, Travis Thomas"
date: "04/08/2026"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, 
                      message = FALSE, results = 'hide')
```

***1. Identification & Cleaning of Variables***

This section identifies key variables with data quality issues and outlines the cleaning techniques applied to improve the dataset. As part of the initial data preparation, the dataset was subsetted to include only swimming events and unnecessary variables such as Season, City, and Games were removed to simplify the analysis. In addition to variable-level cleaning, dataset merging was also performed to enrich the dataset with GDP and population information.


```{r dataReading,echo=FALSE}
setwd("E:/STAT3355")
library(ggplot2)
library(knitr)
olympics_dataset <- read.csv("E:/STAT3355/Athletes_summer_games.csv")
gdp_dataset <- read.csv("E:/STAT3355/gdp_dataset.csv")
population_dataset <- read.csv("E:/STAT3355/Population_dataset.csv", skip = 4)


swimming_olympics <- subset(olympics_dataset, Sport == "Swimming", select = c(-Season, 
                                                                              -City, -Games))

swimming_olympics <- swimming_olympics[swimming_olympics$Year >= 1992, ]
```

**1.1 Age**

The Age variable contained missing values as well as unrealistic observations that fall outside the expected range for Olympic swimmers. Such values can distort statistical summaries and reduce the reliability of the analysis. In addition to that the age variable was also changed to distributed age range, making it convenient for analyzing. 


```{r ageRange, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
summary(swimming_olympics$Age)

swimming_olympics$Age_Group <- cut( swimming_olympics$Age,
                                    breaks = c(0, 18, 25, 35, Inf),
                                    labels = c("Under 18", "18-25", "26-35", "35+"))

summary(swimming_olympics$Age_Group)

```

The distribution of Age before cleaning is shown below.

```{r AgeplotBeforeCleaning, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(swimming_olympics, aes(x = Age)) +
  geom_boxplot(fill = "#2E75B6") +
  labs(title = "Age Distribution BEFORE Outlier Removal") +
  theme_minimal()
```

To improve data quality, all missing values were removed and the dataset was restricted to athletes aged between 13 and 35 years.

```{r AgeplotAfterCleaning, echo=FALSE}
swimming_olympics <- swimming_olympics[
  !is.na(swimming_olympics$Age) &
    swimming_olympics$Age >= 13 & swimming_olympics$Age <= 35, ]

ggplot(swimming_olympics, aes(x = Age)) +
  geom_boxplot(fill = "#1a9850") +
  labs(title = "Age Distribution AFTER Outlier Removal") +
  theme_minimal()

```

This ensures that only realistic observations are retained, improving the accuracy and validity of the dataset.

***1.2 Medal***

The Medal variable included missing values and empty strings, making it difficult to distinguish between athletes who did not win a medal and those with incomplete data. Before the Medal variable was cleaned the variable summary looked as below.

```{r medalSummary, results='markup'}
medal_before <- as.data.frame(table(swimming_olympics$Medal, useNA = "ifany"))
colnames(medal_before) <- c("Medal", "Count")
medal_before$Medal <- ifelse(medal_before$Medal == "", "(blank/missing)", 
                             as.character(medal_before$Medal))
kable(medal_before, caption = "Medal Distribution Before Cleaning")

```

To resolve this issue, all missing and blank entries were replaced with the label "No Medal", and the original variable was removed.

```{r medalCleaned, results='markup'}

swimming_olympics$Medal_Cleaned <- ifelse(is.na(swimming_olympics$Medal) | 
                                  swimming_olympics$Medal == "", "No Medal", 
                                  swimming_olympics$Medal)
medal_after <- as.data.frame(table(swimming_olympics$Medal_Cleaned, useNA = "ifany"))
colnames(medal_after) <- c("Medal", "Count")
kable(medal_after, caption = "Medal Distribution After Cleaning")


```

This transformation ensures a complete and interpretable categorical variable without losing observations.

***1.3 Country Data Integration ***

The original athlete dataset identified countries only by a three-letter NOC code, which lacked demographic context such as population size. To allow for a more comprehensive analysis, the dataset was integrated with a World Bank population dataset. This process involved standardizing the "Country Code" column to match the "NOC" format and performing a left join to align population figures with each athlete's record

```{r CountryplotBeforeCleaning, results='markup'}

kable(head(swimming_olympics[, c("Name", "NOC", "Year")]), 
      caption = "Athlete data prior to merging with population statistics")

```

As shown in the table above, the dataset initially contained only the NOC code to identify each athlete's country, with no demographic information available for further analysis.
To integrate population data, the Country Code column in the World Bank dataset was first renamed to "NOC" to match the identifier used in the athlete dataset. A left join was then performed using the NOC code as the common key, ensuring that all athlete records were retained even if a matching population entry was not found.

```{r CountryplotClean, results='markup'}


colnames(population_dataset)[colnames(population_dataset) == "Country.Code"] <- "NOC"

swimming_olympics <- merge(swimming_olympics, 
                           population_dataset[, c("NOC", "X2016")], 
                           by = "NOC", all.x = TRUE)

colnames(swimming_olympics)[colnames(swimming_olympics) == "X2016"] <- "Population"

kable(head(swimming_olympics[, c("Name", "NOC", "Year", "Population")]),
      caption = "Athlete data after merging with population statistics")
```

The resulting dataset now includes a Population column alongside each athlete record, enabling demographic comparisons across countries. This merging step significantly enriches the dataset and supports more meaningful analysis of how country size or development may relate to Olympic swimming performance.
*note that data for the country "AHO" (Netherlands Antilles) because the country no longer exists

***2. Summary of Data Before Cleaning***

Prior to the cleaning process, the dataset contained several structural and qualitative issues that would have skewed any analytical findings. The original data included all Olympic sports and a vast historical range, creating a high volume of irrelevant data for a swimming-specific study. In terms of specific variables, the Age column was plagued by missing values and extreme outliers—unrealistic ages for elite swimmers—that would have biased mean and median calculations. Similarly, the Medal column contained both NA values and empty strings, making it impossible to distinguish between a "non-medal" finish and a missing data point. Finally, countries were identified only by three-letter NOC codes, providing no socio-economic or demographic context, such as population size, to explain performance variations.

***3. Summary of Data After Cleaning***

The application of targeted cleaning techniques resulted in a standardized and enriched dataset. The data was subsetted to swimming events post-1992 and streamlined by removing redundant variables such as Season, City, and Games. To refine the demographics, missing values in the Age column were removed, and the data was capped at a realistic range of 13 to 35 years. A new categorical variable, Age_Group, was also added to facilitate cohort analysis. The categorical data for medals was standardized by replacing all blanks and NA entries with a clear "No Medal" label, ensuring 100% data completion for this variable. Lastly, by performing a left join with World Bank datasets, the NOC codes were linked to actual population figures, enabling advanced analysis such as calculating medals per capita and understanding the impact of a nation’s size on its Olympic success.

***4. Documentation & Justification ***

The data cleaning process was designed to improve the overall quality, consistency, and usability of the dataset. Removing unrealistic Age values ensures that the dataset accurately reflects the population of Olympic swimmers and prevents distortion in statistical analysis.Replacing missing Medal values with "No Medal" preserves all observations while allowing clear distinctions between medal winners and non-winners. 

The merging of GDP and population datasets adds valuable socioeconomic context, enabling more comprehensive analysis. However, this process introduced missing values, which were addressed by removing incomplete records to maintain data integrity.

Overall, these steps improve the dataset in terms of accuracy, completeness, consistency, and analytical usefulness, ensuring that it is well-prepared for further statistical analysis and visualization.

