---
title: "Data Cleaning Document"
output: pdf_document
author: "Archi Patel,Amanda Contreras, "
date: "04/08/2026"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, 
                      message = FALSE, results = 'hide')
```

***1. Identification & Cleaning of Variables***

This section identifies key variables with data quality issues and outlines the cleaning techniques applied to improve the dataset. As part of the initial data preparation, the dataset was subsetted to include only swimming events and unnecessary variables such as Season, City, and Games were removed to simplify the analysis. In addition to variable-level cleaning, dataset merging was also performed to enrich the dataset with GDP and population information.


```{r dataReading,echo=FALSE}
setwd("E:/STAT3355")
library(ggplot2)
library(knitr)
olympics_dataset <- read.csv("E:/STAT3355/Athletes_summer_games.csv")
gdp_dataset <- read.csv("E:/STAT3355/gdp_dataset.csv")
population_dataset <- read.csv("E:/STAT3355/Population_dataset.csv", skip = 4)


swimming_olympics <- subset(olympics_dataset, Sport == "Swimming", select = c(-Season, 
                                                                              -City, -Games))

swimming_olympics <- swimming_olympics[swimming_olympics$Year >= 1992, ]
```

**1.1 Age**

The Age variable contained missing values as well as unrealistic observations that fall outside the expected range for Olympic swimmers. Such values can distort statistical summaries and reduce the reliability of the analysis. In addition to that the age variable was also changed to distributed age range, making it convenient for analyzing. 


```{r ageRange, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
summary(swimming_olympics$Age)

swimming_olympics$Age_Group <- cut( swimming_olympics$Age,
                                    breaks = c(0, 18, 25, 35, Inf),
                                    labels = c("Under 18", "18-25", "26-35", "35+"))

summary(swimming_olympics$Age_Group)

```

The distribution of Age before cleaning is shown below.

```{r AgeplotBeforeCleaning, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(swimming_olympics, aes(x = Age)) +
  geom_boxplot(fill = "#2E75B6") +
  labs(title = "Age Distribution BEFORE Outlier Removal") +
  theme_minimal()
```

To improve data quality, all missing values were removed and the dataset was restricted to athletes aged between 13 and 35 years.

```{r AgeplotAfterCleaning, echo=FALSE}
swimming_olympics <- swimming_olympics[
  !is.na(swimming_olympics$Age) &
    swimming_olympics$Age >= 13 & swimming_olympics$Age <= 35, ]

ggplot(swimming_olympics, aes(x = Age)) +
  geom_boxplot(fill = "#1a9850") +
  labs(title = "Age Distribution AFTER Outlier Removal") +
  theme_minimal()

```

This ensures that only realistic observations are retained, improving the accuracy and validity of the dataset.

***1.2 Medal***

The Medal variable included missing values and empty strings, making it difficult to distinguish between athletes who did not win a medal and those with incomplete data. Before the Medal variable was cleaned the variable summary looked as below.

```{r medalSummary, results='markup'}
medal_before <- as.data.frame(table(swimming_olympics$Medal, useNA = "ifany"))
colnames(medal_before) <- c("Medal", "Count")
medal_before$Medal <- ifelse(medal_before$Medal == "", "(blank/missing)", 
                             as.character(medal_before$Medal))
kable(medal_before, caption = "Medal Distribution Before Cleaning")

```

To resolve this issue, all missing and blank entries were replaced with the label "No Medal", and the original variable was removed.

```{r medalCleaned, results='markup'}

swimming_olympics$Medal_Cleaned <- ifelse(is.na(swimming_olympics$Medal) | 
                                  swimming_olympics$Medal == "", "No Medal", 
                                  swimming_olympics$Medal)
medal_after <- as.data.frame(table(swimming_olympics$Medal_Cleaned, useNA = "ifany"))
colnames(medal_after) <- c("Medal", "Count")
kable(medal_after, caption = "Medal Distribution After Cleaning")


```

This transformation ensures a complete and interpretable categorical variable without losing observations.

***1.3 Country Data Integration ***

The original athlete dataset identified countries only by a three-letter NOC code, which lacked demographic context such as population size. To allow for a more comprehensive analysis, the dataset was integrated with a World Bank population dataset. This process involved standardizing the "Country Code" column to match the "NOC" format and performing a left join to align population figures with each athlete's record

```{r CountryplotBeforeCleaning, results='markup'}

kable(head(swimming_olympics[, c("Name", "NOC", "Year")]), 
      caption = "Athlete data prior to merging with population statistics")

```

As shown in the table above, the dataset initially contained only the NOC code to identify each athlete's country, with no demographic information available for further analysis.
To integrate population data, the Country Code column in the World Bank dataset was first renamed to "NOC" to match the identifier used in the athlete dataset. A left join was then performed using the NOC code as the common key, ensuring that all athlete records were retained even if a matching population entry was not found.

```{r CountryplotClean, results='markup'}


colnames(population_dataset)[colnames(population_dataset) == "Country.Code"] <- "NOC"

swimming_olympics <- merge(swimming_olympics, 
                           population_dataset[, c("NOC", "X2016")], 
                           by = "NOC", all.x = TRUE)

colnames(swimming_olympics)[colnames(swimming_olympics) == "X2016"] <- "Population"

kable(head(swimming_olympics[, c("Name", "NOC", "Year", "Population")]),
      caption = "Athlete data after merging with population statistics")
```

The resulting dataset now includes a Population column alongside each athlete record, enabling demographic comparisons across countries. This merging step significantly enriches the dataset and supports more meaningful analysis of how country size or development may relate to Olympic swimming performance.
